---
title: "Check of raw data in IDS"
author: "Göran Broström"
date: "December 22, 2016"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE, cache = FALSE}
library(dplyr)
library(knitr)
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

## Data from scratch

First we read data, from IDS and from the created data file.
```{r readin}
load("ef.rda")
source("R/read_IDS.R")
individual <- read_IDS()
```

Then we check the amount of information for each individual in IDS, present in *ef*.
```{r checkin}
id <- unique(ef$Id_I)
indiv <- individual[individual$Id_I %in% id, ]
xx <- with(indiv, tapply(Id_I, Id_I, length))
table(xx)
```
OK, this is a little difficult to interpret, let's look at them in *skum*.

```{r inskum}
library(skum)
oj <- obs[obs$id %in% id, ]
length(unique(oj$id))
dim(oj)
```
Hmm, no connection between *id* in *observation* and *Id_I* in IDS-data. We have to go back to IDS.
We start by noting that rows without any date can safely be discarded.

```{r dateweq}
oo <- rowSums(indiv[, 8:16])
indiv <- indiv[oo > 0.5, ]
xx <- with(indiv, tapply(Id_I, Id_I, length))
table(xx)
```

This tells us that four individuals had three notations in the IDS (modified), nine had four, etc. The guys with three (selected columns):

```{r guysthree}
idi <- names(xx[xx == 3])
indiv[indiv$Id_I %in% idi, c("Id_I", "Type", "Value", "Day", "Month", "Year")]
```

Obviously only "birth" information. In contrast, the guy with 170 notations (watch up, several pages!):

```{r guysmany}
idi <- names(xx[xx == 170])
indiv[indiv$Id_I %in% idi, c("Id_I", "Type", "Value", "Day", "Month", "Year")]
```

Almost 80 years of information $\ldots$. What are the different possible values of *Type*?

```{r type}
sort(unique(indiv$Type))
```


## Dates

Let's fix dates in more readable form.

```{r dates}
datum <- with(indiv, paste(Year, Month, Day, sep = "-"))
indiv$datum <- as.Date(datum, format = "%Y-%m-%d")
datum <- with(indiv, paste(Start_year, Start_month, Start_day, sep = "-"))
indiv$enter <- as.Date(datum, format = "%Y-%m-%d")
datum <- with(indiv, paste(End_year, End_month, End_day, sep = "-"))
indiv$exit <- as.Date(datum, format = "%Y-%m-%d")
indiv <- select(indiv, Id_I, Type, Value, datum, enter, exit)
indiv <- filter(indiv, !(Type %in% c("BAPTISM_LOCATION", "CHILDBIRTH_ASSISTANT",
                                     "FUNERAL_DATE", "FUNERAL_LOCATION",
                                     "MARRIAGE_LOCATION", "VACCINATION")))
indiv <- filter(indiv, !(Type %in% c("OCCUPATION", "OCCUPATION_HISCO")))
kable(head(indiv))
anydate <- indiv$datum
omiss <- is.na(anydate)
anydate[omiss] <- indiv$exit[omiss]
indiv$anydate <- anydate
rm(anydate)
length(unique(indiv$Id_I))
indiv <- filter(indiv, !is.na(anydate))
length(unique(indiv$Id_I))
```

We remove rows with *Type == "BAPTISM_DATE", "BIRTH_LOCATION", "DEATH_LOCATION*

```{r removemore}
indiv <- filter(indiv, !(Type %in% c("BAPTISM_DATE", "BIRTH_LOCATION", "DEATH_LOCATION")))
with(indiv, table(Type))
```

Now we check the min and max date for each person.

```{r minmax}
mind <- with(indiv, tapply(anydate, Id_I, min))
maxd <- with(indiv, tapply(anydate, Id_I, max))
wox <- data.frame(id = names(maxd),
                  first = as.Date(mind, origin = "1970-01-01"), 
                  last = as.Date(maxd, origin = "1970-01-01"),
                  age = (maxd - mind) / 365.2425)
wox$birthYear <- wox$first %>% 
    as.character() %>% 
    substr(1, 4) %>% 
    as.numeric()
id <- indiv$Id_I[indiv$Type == "DEATH_DATE"]
##wox$death <- FALSE
wox$death <- wox$id %in% id
```

