---
title: "Description of the Skellefteå data"
author: "Göran Broström"
date: "August 18, 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

A closer look at the data created from *Episodes_file_creator*. Start by loading it
```{r loaddata}
load("ef.rda")
ef$event <- as.logical(ef$event)
summary(ef[, c("Child_sex", "G0_Births", "G0_InfD_cat", "Mother_birthdate", "enter", "exit", "event")])
ef1 <- ef[!duplicated(ef$MotherID), ]
```

# Grandmothers, mothers and mother-siblings

How many mothers and grandmothers?

```{r howmany}
nrow(ef1) # No. of mothers
length(unique(ef1$GrandmotherID)) # No. of grandmothers
table(table(ef1$GrandmotherID)) # 'Mother siblings'
```

How many grandmothers are also mothers (and vice versa)?

```{r countgm}
with(ef1, sum(unique(GrandmotherID) %in% MotherID)) 
with(ef1, sum(MotherID %in% GrandmotherID)) 
```
(Of course!)

This frequent reappearance of women in dual roles and also the sibling frequency among mothers create dependency problems.
Two ways to deal with it:

1. Restrict data.

2. Ignore it.