--- 
title: "Models for intergenerational transmissions in infant mortality"
author: "Göran Broström"
date: "August 15, 2016"
output: pdf_document
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE) 
options(show.signif.stars = FALSE, digits = 5)
```

# The data

There are data on births and infant deaths for mothers and their grandmothers.
More specifically, we have, for each mother, the following information:

* n: Number of births 

* $B_1, \ldots, B_n$: Age of mother at child births.

* $T_1, \ldots, T_n$: Life lengths of the born kids. Censored at the age of one year, numbered according to birth order. 

* $D_1, \dots, D_n$: Indicators of $T_j < 1, \; j = 1, \ldots, n$. 

* $Z$: The number of infant deaths facing the corresponding grandmother. 

* $n_Z$: Grandmother's number of births. 

* $\mathbf{x}$: A vector of covariates.

The explanatory variable $(Z, n_Z)$ is of primary interest. And since we are
interested of its effect on "infant mortality", it is natural to restrict
attention to $D_1, \ldots, D_n$ as response, but it is also possible to
complicate things and try to model the complete data set.

# A binary model

If we assume that, given the mother, $T_1, \ldots, T_n$ are independent random
variables with survivor function $S_j$, then we have that $D_1, \ldots, D_n$ are
independent Bernoulli random variables with

$$ P(D_j = 1) = P(T_j < 1) = 1 - S_j(1), \quad j = 1, \ldots, n. $$

Assuming that the relation between these survivor functions follow a
*proportional hazards* model, i.e.,

$$ S_j(1) = q^{k_j}, \quad j = 1, \ldots, n, $$ for some real number $0 < q <
1$, then, assuming proportional hazards relations visavi all covariates, we get 
$$ P(Z, \mathbf{x}, j) = P(T_j < 1; Z, x) = 1 - \exp(-e^{\alpha + \beta g(Z,
n_Z) + \boldsymbol{\gamma} \mathbf{x}}), $$ where $g$ is a suitable function,
for instance $$ g(Z, n_Z) = \frac{Z}{n_Z} = \bar{Z}, $$ and the parity
information is absorbed into $\mathbf{x}$. The likelihood contribution from one
mother (number $i$, say) is then

$$ L_i(\alpha, \beta, \boldsymbol{\gamma} \mid \mathbf{D}_i) = \prod_{j=1}^n
\bigl\{1 - \exp(-e^{\alpha + \beta g(Z, n_Z) + \boldsymbol{\gamma}
\mathbf{x}})\bigr\}^{D_i} \bigl\{\exp(-e^{\alpha + \beta g(Z, n_Z) +
\boldsymbol{\gamma} \mathbf{x}})\bigr\}^{1-D_i} $$

We recognize this as the likelihood kernel of a *Bernuolli* model with the
*cloglog* link function, that is, *logistic regression* with a cloglog link.
There one problem with this, though: With data from only one mother (as above),
the model is not identifiable, because the variable $Z$ is constant over the
mother. We can estimate $\alpha + \beta g(Z, n_Z)$, but not isolate the effect
of $(Z, n_Z)$. With data from many mothers we can solve this dilemma in two
ways: (i) Let $\alpha$ be constant over mothers, or (ii) fit a *random effects
model*, where $\alpha_1, \ldots, \alpha_N$ are regarded as a random sample from
som suitable distribution with finite variance. Method (i) is clearly
unsatisfactory (would require an assumption about total independence), but (ii)
is possibe without strong assumptions.

## A *simple* binary model

A very simple and straightforward model, that requires no restrictive
assumptions, is to skip all covariates and have as explanatory variable

$$ R = g(Z, n_Z) = I(Z > 0) $$ and response $$ Y = I\biggl(\sum_{j=1}^n D_j >
0\biggr) $$

where $I(A)$ is the indicator function of the event $A$, equal to unity if $A$
is *TRUE*, zero otherwise. The analysis will then boil down to a $\chi^2$ test
in a $2\times 2$ table, see my blog post
[Inheritance](http://capa.ddb.umu.se/rr/jekyll/rstats/2016/06/01/inheritance.html)
for an example.


# A survival analysis approach

In this approach we utilize the full data set $(B_i, T_i, D_i, Z_i, X_i), i = 1, \ldots, n$ and
a proportional hazards assumption, as in the binary approach. The research
question is then more indirect, e.g.: "Does grandmother's infant deaths affect the
life length of mother's children, restricted to the first year of life?"

The same identifiability problem as above shows here, and it can be handled in
essentially the same way, that is (ii) is the preferred way.

# Summing up

So, we have (at least) three possible models to our disposal; which one should we choose?
My maxim is *the simpler, the better*, which would point to the $2\times 2$ table above, but it is clearly OK to apply one of the more elaborate models, if one is willing to investigate the necessary model checks (*proportional hazards*, *conditional independence*, etc.).

But, whatever our choice is, I recommend employing the simple model(s) as a sanity check: If it shows a  completely different result, we should be worried.

# Application

I exemplify with a subset of the data from Skellefteå. It looks like this (the first ten lines):

```{r datamoth, echo = FALSE}
load("ef.rda")
dat <- data.frame(m.id = ef$MotherID,
                  m.birthdate = ef$Mother_birthdate,
                  enter = ef$enter,
                  exit = ef$exit,
                  event = ef$event,
                  sex = ef$Child_sex,
                  parity = ef$Birth_order,
                  gm.id = ef$GrandmotherID,
                  gm.births = ef$G0_F_Births + ef$G0_M_Births,
                  gm.m.births = ef$G0_M_Births,
                  gm.deaths = ef$G0_F_InfD + ef$G0_M_InfD,
                  gm.m.deaths = ef$G0_M_InfD)
dat <- dat[dat$gm.births > 1 & dat$gm.m.births > 0, ]
dat <- dat[order(dat$m.id, dat$enter), ]
rownames(dat) <- 1:NROW(dat)
dat$gm.deaths.cut <- cut(dat$gm.deaths, c(-1, 0.5, 1.5, 25), labels = c("0", "1", "2+"))
dat$m.birthdate <- (as.numeric(dat$m.birthdate) - 51478) / 3650.2425 
knitr::kable(head(dat[, c("m.id", "m.birthdate", "parity", "enter", "exit", "event", "gm.deaths")], 10))
```

*m.birthdate* is a centered version of *Mother's birthdate*, *enter* is mother's age at delivery, and *exit* is the age of mother at the death of the infant if it happens before age 1, $enter + 1$ otherwise. *event* indicates an infant death.

## The binary model

In the full binary model, the variable *event* is the response in a logistic regression, with clustering on *mother's ID*.

```{r binclust, cache = TRUE}
library(lme4)
fit <- glmer(event ~ gm.deaths.cut + m.birthdate + (1 | m.id), data = dat, 
             family = binomial(link =  cloglog))
summary(fit)
```

Note the warning: *The model is nearly unidentifiable*. This is not surprising, since we *cluster on mother* and the *explanatory variable is constant over mothers*! This is not a good way to take into account the "natural" biological/social variation between mothers. A better approach is to try to find a summary measure for each mother, for instance the total number of infant deaths. Which leads to

## The *simple* (aggregated) binary model

Aggregating over mothers gives

```{r aggregmot}
indx <- tapply(dat$event, dat$m.id)
deaths <- tapply(dat$event, dat$m.id, sum)
dat$deaths <- deaths[indx]
expos <- tapply(dat$exit - dat$enter, dat$m.id, sum)
dat$expos <- expos[indx]
dat1 <- dat[dat$parity == 1, ]
rownames(dat1) <- 1:NROW(dat1)
knitr::kable(head(dat1[, c("gm.births", "gm.deaths", "deaths", "expos", "m.birthdate")]))
```

A Poisson regression with *deaths* as response and *expos* as offset (log scale):

```{r poiss}
fit <- glm(deaths ~ offset(log(expos)) + gm.deaths.cut + m.birthdate, data = dat1, family = poisson)
summary(fit)
```

Note the similarity of the estimates with the "full" binary model! And note further that here we do not have to worry about "random effects" (but see the Conclusions).

Note further the bad/erroneous presentation of $p$-values! Unfortunalely, standard output from statistical software looks like this (even from R! But in R I can do something about it ...). The problem is twofold: (i) We do not want $p$-values for *levels* of a variable, but for the variable itself. (ii) The $p$-values presented are of the so-called "Wald" type, which is perfectly OK in linear models but can be a *disaster in non-linear models*. The method of choice is to use $p$-values based on the *likelihood ratio test* (LRT).
In R, you get them by applying the *drop1* function to the output (*fit*) from the regression:

```{r droppois}
drop1(fit, test = "Chisq")
```

Note the difference; unfortunately you need to do some manual work to get it properly into the regression table. I have written some code in R that automates the procedure for $\text{\LaTeX}$ output, but not (yet) for *R markdown*, which I use here.

## Survival analysis

Walking in the other direction, that is, making the model *more* complex and fragile, we can apply survival analysis (Cox regression).

```{r coxreg}
library(eha)
fit <- coxreg(Surv(exit - enter, event) ~ gm.deaths.cut + m.birthdate, data = dat)
dr <- drop1(fit, test = "Chisq")
summary(fit)
dr
```

No big difference in results from the previous attempts. However, here we have to worry about the possibility of "frailty" ("random effects", etc) due to dependency of sibling fates.

A check of proportionality by *gm.deaths*, another worry:

```{r checkprop}
fit <- coxreg(Surv(exit - enter, event) ~ strata(gm.deaths.cut) + m.birthdate, data = dat)
plot(fit, col = c("black", "blue", "red"), xlab = "Age", ylab = "Cumulative hazards")
```

A *clear non-proportionality*: It is mothers with exactly one death by grandmother who deviate. Initially, up to two months, the risk is the same as for mothers with heavier grandmother baggage, but after surviving the first two months, the rest is easy sailing, even rosier than for those mothers whose grandmother didn't suffer any fatality.

# Conclusions

Based on these small experiments, I would recommend using the *simple binary model*, that is, *Poisson regression*, as above. It is the model that requires the least assumptions: The survival analysis has two severe problems: (i) individual frailty, and (ii) non-proportionality on the important explanatory variable (in our data). On the other hand, (ii) opens up for new interesting research questions, about varying effects over the span of the first year of life.

Another "dependency problem" that is not addressed anywhere(?): The same individual may pop up in our data set as a grandmother *and* as a mother, even as a *child*. This could be adressed by restricting the presence of each individual: only one performance! This would also exclude siblings to be present as mothers.

Pick your poison.