---
title: Clustering and intergenerational transfers of infant mortality in 19th century
  northern Sweden
author: "Elisabeth Engberg, Sören Edvinsson, and Göran Broström"
date: "August 21, 2016"
output:
  pdf_document: 
    fig_height: 4
    fig_width: 5
    fig_caption: yes
  html_document: default
  word_document:
    fig_caption: yes
    fig_height: 4
    fig_width: 5
    reference_docx: word-styles-reference-01.docx
bibliography: cluster.bib
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, comment = NA)
options(show.signif.stars = FALSE)
library(knitr)
library(eha)
```
# Introduction

# Area

 The area under study is the Skellefteå region on the Gulf of Bothnia. 
The analyses of the Skellefteå region cover the period 1831–1900.

The Skellefteå region consists of the parish of Skellefteå and the parish of
Byske. Byske was a part of Skellefteå parish until 1875. For the remaining
part of the century it was an independent parish. Two other parishes which
became independent in the early nineteenth century, Jörn and Norsjö, were
incorporated into the parish of Skellefteå for a number of years at the be-
ginning of the century and are included in the analyses for these years. The
region is situated in the province of Västerbotten in the northern part of
Sweden on the Gulf of Bothnia. The parish was one of the largest parishes in
Sweden both in area and in population during the nineteenth century. The
population began to increase slowly during the seventeenth and eighteenth
centuries. By the turn of the nineteenth century, about 6 900 inhabitants
lived in the parish. The population of the parish increased rapidly during the
first half of the century, reaching approximately 14 000 by 1850. By 1900 the
two parishes had nearly 30 000 inhabitants [@as94]. The increase
in population was mainly the result of a high natural growth. Mortality was
comparatively low. Fertility was high, not only by Swedish standards, but
also in an international comparison and there are no indications of family
planning. Total fertility fluctuated around five children per woman and, al-
though fertility did decline during the nineteenth century, the actual fertility
transition occurred late in the district [@as94]. The rate of 
illegitimacy was low in comparison with many other parts of Northern Sweden
where frequent pre-nuptial conceptions and illegitimate births were common.
The illegitimacy rate fluctuated between three and six per cent during the
nineteenth century [@as94]. As the population increased new land
was cultivated and, although some villages became quite large, population
density, on the whole, was low.

During most of the nineteenth century industrialisation had little impact
on the local economy. Some small sawmills were established early in the
century. The majority of the population (in 1835 approximately 85 per cent)
made their living from farming, mainly animal husbandry. The farmers were
mainly freeholders and there were no large estates in the region. The distri-
bution of economic resources among this group was comparatively equal.


# Data sources

The analysis is based on information found in the Swedish church book
records, that is, birth, death, migration, and catechetical registers. In the
catechetical registers the clergy kept a continuous record of all demographic
events for all individuals residing in a parish, making it possible to follow in-
dividuals over time and to identify their relatives. Furthermore, information
on literacy and occupation was recorded, including any changes in status. A
selected number of parishes are digitalized by the [Demographic Data Base at
Umeå University](http://www.cedar.umu.se/). The entries in the catechetical
registers have been linked for each individual making possible longitudinal
analyses.

# Implementation of IDS

A closer look at the data created from *Episodes_file_creator*. It contains one record per birth and is used for the suvival analysis approach, where the influence of grandmother's infant deaths on her grandchild's survival chances is studied.

# Results

## Descriptive statistics

```{r loaddata}
load("ef.rda")
ef$event <- as.logical(ef$event)
levels(ef$G0_InfD_cat) <- c("0", "1", "2+", "2+")
ef$M_birthdate <- as.numeric(ef$Mother_birthdate) / 365.2425 + 145 # 1825 == 0!
ef <- ef[ef$M_birthdate + 1825 > 1800, ] ## NOTE !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
ef$parity <- cut(ef$Birth_order, c(0.5, 1.5, 2.5, 5.5, 25), labels = c("1", "2", "3-5", "6+"))
##kable(summary(ef[, c("Child_sex", "Mother_birthdate", "M_birthdate", "enter", "exit", "event")]), type = "pandoc")
```

This data frame is then aggregated to one record per mother for the Poisson regression approach, where  the influence of grandmother's infant deaths on the number of her daughter's' infant deaths is studied.

### The covered time period

See Figure 1 for the distribution of mothers' birth dates.

```{r aggrfile}
indx <- tapply(ef$MotherID, ef$MotherID)
ef$M_Births <- with(ef, tapply(MotherID, MotherID, length)[indx])
ef$M_Deaths <- with(ef, tapply(event, MotherID, sum)[indx])
ef$M_expon <- with(ef, tapply(exit - enter, MotherID, sum)[indx])
ef1 <- ef[!duplicated(ef$MotherID), ]
##kable(summary(ef1[, c("M_Births", "M_Deaths", "G0_Births", "G0_InfD_cat")]))
```

```{r sample, fig.cap = "Distribution of mother birthdates."}
plot(density(ef1$M_birthdate + 1825), xlab = "Year", main = "", col = "blue")
abline(h = 0)
```

### Infant mortality by year

```{r imryear}

```


### Grandmothers, mothers and mother-sisters


```{r howmany}
n.m <- nrow(ef1) # No. of mothers
n.gm <- length(unique(ef1$GrandmotherID)) # No. of grandmothers
x <- table(table(ef1$GrandmotherID)) # 'Mother siblings'
```

There are `r n.m` mothers and `r n.gm` grandmothers in the data, so obviously there are many sister groups in the data. The distribution of the sizes of sibling groups is shown in Figure 2.

```{r sibmoth, fig.cap = "Distribution of sister group sizes."}
barplot(x, xlab = "Size", ylab = "Frequency")
```


How many grandmothers are also mothers (and vice versa)?

```{r countgm}
m.gm <- with(ef1, sum(unique(GrandmotherID) %in% MotherID)) 
```

The answer is `r m.gm`, or `r round(m.gm / n.gm * 100, 2)` per cent of the grandmothers.

## Standard results

### Poisson regression

The expected value of the number of infant deaths $D_i$ for mother No. $i, \, i = 1, \ldots, n$, is modeled by a Poisson distribution as 

$$
E(D_i) = R_i e^{\boldsymbol{\beta} \mathbf{x}_i},
$$
where $R_i$ is total risktime for mother No. $i$, $\mathbf{x}_i$ a vector of her explanatory variables, and $\boldsymbol{\beta}_i$ is the vector of regression coefficients. (For a mother with no infant deaths, the risktime is equal to her number of births.)

Formally, $R_i$ is entered into the model as an *offset* after taking logs. The logged value is stored in the variable *expon*. The call to the **R** [@R] function *glm* is

```{r callglm, echo = TRUE}
fit.p <- glm(M_Deaths ~ offset(log(M_expon)) + 
               G0_InfD_cat + G0_Births + M_birthdate, 
             data = ef1, family = poisson)
```
The result of this call is presented in two steps: First, *the statistical significance* is calculated and shown, in Table 1, and then the effect sizes are graphically evaluated, in Figure 1.


```{r poisson, results = 'asis'}
options(digits = 8)
fit.p <- glm(M_Deaths ~ offset(log(M_expon)) + G0_InfD_cat + G0_Births + M_birthdate, data = ef1, family = poisson)
dr <- drop1(fit.p, test = "Chisq")

kable(dr, format = "pandoc", caption = "Analysis of deviance (likelihood ratio tests). Poisson regression.", digits = 5)
x <- summary(fit.p)$coefficients
##outp <- matrix(0, ncol = 4, nrow = 6)
##outp[, 1] <- c(x[1, 1], NA, x[, 1][-1])
##outp[, 2] <- c(x[1, 2], NA, x[, 2][-1])
##outp[, 3] <- exp(outp[, 1])
##outp[, 4] <- c(dr[[5]][1:2], NA, NA, dr[[5]][3], dr[[5]][4])
##rownames(outp) <- c("(intercept)  ", "G0_InfD   0", "          1", "          2+", "G0_Births", ##"Mother_Birthdate")
##colnames(outp) <- c("Coef", "S.E.", "exp(Coef)", "L-R p")
##knitr::kable(outp, format = "pandoc", caption = "Table 2. Poisson regression.", label = "tab:tab1")
##outp
##print(xtable(outp), type = "html")
##summary(fit)
```

So, the likelihood ratio test (LRT) shows that *G0_InfD_cat* is highly *statistically* significant. 

Is it also *practically* significant? Figure 1 shows the *relative* risk increase induced by exactly one infant death by grandmother (GM) and at least two, compared to cases with no infant deaths (baseline). 

```{r poiseff, fig.cap = "Relative risk increase of infant mortality, compared to no grandmother infant deaths.", fig.width = 4}
cat0 <- exp(fit.p$coef[1] + 7 * fit.p$coef[5])
cat1 <- exp(sum(fit.p$coef[1:2]) + 7 * fit.p$coef[5])
cat2 <- exp(sum(fit.p$coef[c(1, 3)]) + 7 * fit.p$coef[5])
barplot(100 * ((c(cat1, cat2) - cat0)/ cat0), names.arg = c("1 GM death", "2+ GM deaths"), 
        ylim = c(0, 25), ylab = "Relative IMR increase (per cent)")
```

The *baseline risk* is `r round(100 * cat0, 2)` per cent. The *relative* risk increase is quite substantial for 2+ grandmother deaths, but the *absolute* increase (`r round(100 * (cat2 - cat0), 2)` per cent) is modest. These calculations are made for a mother born 1825 and her grandmother gave birth to seven children ("average" values). 

### Survival analysis

Here the full data set is used, that is, the infants are the units of analysis. The **R** package *eha* [@eha, @ehar] is used, and the explanatory variables are almost the same as in the Poisson regression analysis. The difference is that instead of mother's birthdate, the infant's birthdate is used (no big difference).

```{r surv}
fit.c <- coxreg(Surv(exit - enter, event) ~ G0_InfD_cat + I(G0_Births - 7) + I(M_birthdate + enter) + parity, data = ef)
dr.c <- drop1(fit.c, test = "Chisq")
kable(dr.c, format = "pandoc", caption = "Analysis of deviance (likelihood ratio tests). Cox regression.", digits = 5)
```

The statistical significance of grandmother's infant deaths is obvious. The estimated relative risks are best shown by plotting the *cumulative hazards* for the three groups.

```{r relcumhaz, fig.cap = "Cumulative hazards by the number of grandmother's infant deaths."}
fit.str <- coxreg(Surv(exit - enter, event) ~ strata(G0_InfD_cat) + I(G0_Births - 7) + I(M_birthdate + enter) + parity, data = ef)
plot(fit.str, col = 1:3, lty = 1:3, xlab = "Age", ylab = "Cumulative hazards.")
```

There is an evident deviation from the assumption of *proportional hazards*: It is cases with exactly one grandmother infant death that deviates. However, this does not disturb the main conclusion: Two or more grandmother infant deaths is harmful, quite in line with the results from the Poisson regression. 

## Extended models

### The Poisson model

It turns out that the model fit to data is somewhat better with *number of births* as the offset (log scale). Also showing a slight improvement is to use *grandmother's IMR* as explanatory variable rather than her *absolute number of deaths*. The IMR is defined as the number of infant deaths divided by the number of births. These changes also implies that the comparison *mother vs. grandmother* happens on a probability scale rather than on the intensity one.

```{r refinedpois, results = "asis"}
ef1$G0_IMR <- ef1$G0_InfD / ef1$G0_Births
fit.p2 <- glm(M_Deaths ~ offset(log(M_expon)) + G0_IMR + M_birthdate + parity, data = ef1, family = poisson)
dr <- drop1(fit.p2, test = "Chisq")
kable(dr, format = "pandoc", caption = "Poisson regression, grandmother's IMR. Analysis of deviance.", digits = 5)
```

The general conclusion is not changed: Grandmother's IMR has a strong influence on infant mortality,  and, indirectly, on her daughter's IMR.

The size of the effect is shown in Figure 3.

```{r plotimr, fig.cap = "Probability of infant death for mother (M_IMP) by probability of infant death for grandmother (GM_IMP). Poisson regression."}
x <- seq(0, 1, length = 1000)
mimr <- with(fit.p2, exp(coefficients[1] + coefficients[2] * x))
plot(x, mimr, type = "l", col = "blue", xlab = "GM_IMP", ylab = "M_IMP", ylim = c(0, 0.17))
abline(h = 0)             
```

### Survival analysis

The same modification as in the Poisson case is introduced here.

```{r extsurv}
ef$G0_IMR <- ef$G0_InfD / ef$G0_Births
fit.c2 <- coxreg(Surv(exit - enter, event) ~ G0_IMR + I(M_birthdate + enter) + parity, data = ef)
dr.c2 <- drop1(fit.c2, test = "Chisq")
kable(dr.c2, format = "pandoc", caption = "Analysis of deviance (likelihood ratio tests). Cox regression, extended model.", digits = 5)
```

# Conclusion

This frequent reappearance of women in dual roles and also the sibling frequency among mothers create dependency problems.
Two ways to deal with it:

1. Restrict data.

2. Ignore it.

# References
